---
title: "COFINFAD: Colombian Fintech Cluster Analysis"
---

## The Objective

The objective of this project is to understand the behaviors behind customer transactions and how socio demographic factors such as age, income, and occupation affect customer transactions. By applying statistical correlation techniques, we aim to identify which personal attributes are the true drivers of financial activity within the Colombian fintech ecosystem.

Customer clustering techniques are used to group customers with similar financial behaviors and socio-demographic characteristics. By identifying distinct customer segments, the analysis provides insights into how different groups interact with fintech services, including their transaction patterns, savings behavior, and digital engagement levels.

## Data Preparation

The required R packages are loaded using the `pacman::p_load()` function, which automatically installs the packages if they are not already available and loads them into the working environment. These libraries support data manipulation, visualization, and clustering analysis required for the visual analytics application.

```{r warning=FALSE}
pacman::p_load(ggrepel, patchwork,
               ggthemes, hrbrthemes,
               tidyverse, cluster,
               factoextra)

```

The COFINFAD dataset consists of two files: a customer-level dataset containing 48,723 customer records and a transaction-level dataset containing over 3.1 million transaction entries. The customer dataset provides demographic, behavioral, engagement, and satisfaction attributes, while the transaction dataset stores transaction-level records including amount, date, and transaction type.

```{r}

# Import datasets
# Import datasets
customer_data <- read_csv("data/customer_data.csv")

transactions_data <- read_csv("data/transactions_data.csv")

# Preview datasets
head(customer_data)
head(transactions_data)
```

```{r}
colnames(customer_data)

```

```{r}
colnames(transactions_data)
```

## Variable Selection for Clustering

The selected variables were chosen to represent multiple dimensions of customer behavior. Variables such as `tx_count`, `avg_tx_value`, `monthly_transaction_count`, and `transaction_frequency` capture transaction intensity, while `active_products`, `app_logins_frequency`, and `feature_usage_diversity` reflect platform engagement. `satisfaction_score`, `nps_score`, and `churn_probability` were included to account for customer sentiment and retention risk, and `customer_lifetime_value` was included to reflect long-term business value.

```{r}
cluster_data <- customer_data %>%
select(
age,
active_products,
app_logins_frequency,
feature_usage_diversity,
tx_count,
avg_tx_value,
monthly_transaction_count,
transaction_frequency,
satisfaction_score,
nps_score,
churn_probability,
customer_lifetime_value
)
```

### Feature Correlation Heatmap

The correlation heatmap was used to examine relationships between selected variables. Highly correlated variables were evaluated to avoid redundancy in clustering.

```{r}
library(corrplot)



cor_matrix <- cor(cluster_data, use = "complete.obs")

corrplot(cor_matrix,
         method = "color",
         type = "upper",
         tl.cex = 0.7)
```

The heatmap shows that several transaction-related variables are positively correlated, especially variables linked to transaction intensity and value. This is expected because customers who transact more frequently often also generate higher total activity. However, the correlations are not uniformly strong across all variables, suggesting that the selected features still capture different aspects of customer behavior and remain suitable for clustering.

## Handling Missing Values

Before performing clustering, missing values were handled to ensure data quality and prevent errors during analysis. Observations containing missing values were removed to create a clean dataset for clustering.

```{r}
cluster_data <- na.omit(cluster_data)
```

## Data Standardization

Since the selected variables are measured on different scales (e.g., transaction volume, satisfaction scores, and login frequency), the data was standardized. Standardization ensures that all variables contribute equally to the clustering process.

```{r}
cluster_scaled <- scale(cluster_data)
```

## Determining the Optimal Number of Clusters

K-means clustering was selected because the chosen variables are primarily continuous numeric measures that describe customer behavior, engagement, and financial activity. After standardization, K-means provides a simple and interpretable way to group customers based on similarity across these dimensions.

```{r}
set.seed(123)

# take a sample for elbow method
cluster_sample <- as.data.frame(cluster_scaled) %>%
  slice_sample(n = 10000)

# compute WSS manually
wss <- numeric(10)

for (i in 1:10) {
  km <- kmeans(cluster_sample, centers = i, nstart = 10)
  wss[i] <- km$tot.withinss
}

# plot elbow curve
plot(1:10, wss, type = "b",
     xlab = "Number of Clusters",
     ylab = "Within-cluster Sum of Squares",
     main = "Elbow Method for Optimal Number of Clusters")

```

The elbow plot suggests that the rate of improvement in within-cluster variation begins to slow after approximately 3 to 4 clusters. This indicates that additional clusters beyond this point may provide diminishing returns in segmentation quality.

### Silhouette Analysis

```{r}
sil_width <- numeric(9)

for (k in 2:10) {
  km <- kmeans(cluster_sample, centers = k, nstart = 10)
  ss <- silhouette(km$cluster, dist(cluster_sample))
  sil_width[k - 1] <- mean(ss[, 3])
}

plot(2:10, sil_width, type = "b",
     xlab = "Number of Clusters",
     ylab = "Average Silhouette Width",
     main = "Silhouette Analysis for Optimal Number of Clusters")
     
```

The silhouette analysis shows that the highest average silhouette width is achieved at 3 clusters, while 4 clusters also produces a relatively strong result. Since 4 clusters provides a more detailed and interpretable segmentation structure while maintaining acceptable cluster quality, it was selected for the final K-means model.

```{r}
set.seed(123)
kmeans_model <- kmeans(cluster_scaled, centers = 4, nstart = 25)
```

## K-Means Clustering

The final clustering model was estimated using K-means with 4 clusters and 25 random starts to improve stability and reduce the chance of converging to a poor local solution.

```{r}
customer_data$cluster <- kmeans_model$cluster
```

## Cluster Distribution

```{r}
table(customer_data$cluster)
```

```{r}
ggplot(customer_data, aes(x = factor(cluster))) +
  geom_bar(fill = "steelblue") +
  labs(
    title = "Customer Distribution by Cluster",
    x = "Cluster",
    y = "Number of Customers"
  )
```

The resulting cluster sizes are moderately imbalanced, with Cluster 2 representing the largest segment of customers. This suggests that one dominant customer profile exists in the fintech platform, while the remaining clusters capture smaller but still meaningful behavioral segments.

## Cluster Visualization

To better understand the separation between clusters, the clustering results were visualized using dimensionality reduction techniques. This visualization provides an overview of how customers are grouped based on their behavioral and financial characteristics.

```{r}
fviz_cluster(kmeans_model, data = cluster_scaled)
```

### Variance Explained by Principal Components

```{r}
pca_model <- prcomp(cluster_scaled)

fviz_eig(pca_model)


```

The PCA-based cluster visualization shows that the identified customer segments are partially separable in lower-dimensional space. Although some overlap remains, the projection indicates that the selected variables provide sufficient structure to distinguish distinct customer groups.

Analyze Cluster Characteristics

To interpret the meaning of each cluster, the average values of key variables were calculated for each cluster. This allows us to identify behavioral patterns and characteristics that distinguish different customer segments.

```{r}
cluster_summary <- customer_data %>%
  group_by(cluster) %>%
  summarise(
    avg_age = mean(age, na.rm = TRUE),
    avg_products = mean(active_products),
    avg_logins = mean(app_logins_frequency),
    avg_transactions = mean(tx_count),
    avg_tx_value = mean(avg_tx_value),
    avg_satisfaction = mean(satisfaction_score),
    avg_nps = mean(nps_score),
    avg_churn = mean(churn_probability),
    avg_clv = mean(customer_lifetime_value)
  )

cluster_summary
```

The cluster summary shows meaningful differences across segments. Cluster 1 stands out for having the highest average transaction value and customer lifetime value, suggesting a high-value segment. Cluster 2 is the largest group but shows lower average transaction intensity and lower financial value, indicating a mainstream but less valuable segment. Cluster 4 stands out for having the highest number of active products and logins, suggesting a highly engaged digital segment. Cluster 3 appears to sit between these groups, with moderate engagement and transaction activity.

## Cluster Profiling

To better compare the characteristics of different clusters, key behavioral and financial variables were visualized across clusters. This helps highlight the differences between customer segments.

```{r}
ggplot(customer_data, aes(x = factor(cluster), y = tx_count)) +
  geom_boxplot(fill = "orange") +
  labs(
    title = "Transaction Count by Cluster",
    x = "Cluster",
    y = "Transaction Count"
  )
```

The clustering results reveal distinct customer segments within the fintech platform. For instance, some clusters represent highly active customers with frequent transactions and high customer lifetime value, while others represent lower engagement users with fewer financial activities and higher churn probability.

```{r}
ggplot(customer_data, aes(x = factor(cluster), y = avg_tx_value)) +
  geom_boxplot(fill = "skyblue") +
  labs(
    title = "Average Transaction Value by Cluster",
    x = "Cluster",
    y = "Average Transaction Value"
  )
```

The profiling plots confirm that clusters differ not only in transaction count but also in financial value and engagement intensity. This suggests that the identified segments are behaviorally meaningful and can be used to distinguish high-value, low-activity, and digitally engaged customer groups.

### Standardized Comparison of Cluster Characteristics

```{r}
library(tidyr)

cluster_profile_long <- cluster_summary %>%
  pivot_longer(-cluster, names_to = "metric", values_to = "value") %>%
  group_by(metric) %>%
  mutate(value_scaled = as.numeric(scale(value)))

ggplot(cluster_profile_long,
       aes(x = metric, y = value_scaled, color = factor(cluster), group = cluster)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  theme_minimal() +
  labs(
    title = "Standardized Comparison of Cluster Characteristics",
    x = "Metric",
    y = "Standardized Value",
    color = "Cluster"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Because the metrics are measured on very different scales, the cluster comparison values were standardized before plotting. This makes it easier to compare relative differences across clusters without the chart being dominated by high-magnitude variables such as customer lifetime value.

### Cluster Membership Composition

```{r}
ggplot(customer_data, aes(x = factor(cluster), fill = income_bracket)) +
  geom_bar(position = "fill") +
  labs(
    title = "Income Bracket Composition Across Clusters",
    x = "Cluster",
    y = "Proportion"
  )
```

The proportional composition chart helps interpret cluster membership by showing how socio-demographic categories are distributed across the identified customer segments. This supports the practical interpretation of clusters beyond purely numeric summary statistics.

### Interpretation of Customer Segments

-   **Cluster 1** appears to represent a high-value segment with the highest average transaction value and customer lifetime value.
-   **Cluster 2** is the largest segment and likely represents mainstream customers with moderate engagement and lower financial value.
-   **Cluster 3** represents a medium-activity segment with moderate transaction and engagement characteristics.
-   **Cluster 4** appears to be a highly engaged digital segment, with the highest number of active products and app logins.

## UI Design for Proposed Shiny Module
